Cultural Transmission and Technological Transitions during the Late Paleolithic in Korea

### 2022-08-28 analysis/paper/paper-doc.qmd
moved discussion and conclusion text to qmd

We explored assemblages that contain multiple stemmed points to examine the temporal patterns with the premise that stemmed points from one assemblage were made by the same group of people. There are four assemblages, three assemblages from the Suyanggae site (SYG) and one from the Yongsandong site (YS) that contain seven or more stemmed points. SYG1_2 and SYG6_2 are from Phase 2 and SYG1_3 and YS_3 are from Phase 3. Following the previous studies of [@bettinger1997evolutionary; @bettinger1999point; @eerkens2008cultural; @mesoudi2008cultural; @garvey2018current], we only compared CV values of attributes, body length, maximum length, mid width, and tang width. @fig-eight shows that CV values for the four assemblages are mostly below 25, which represent transmissions through indirect bias. Among the assemblages, SYG1_2 (n=7) has higher CV values as well as the widest range of confidence intervals. Compared to the Phase 2, CV values of Phase 3 are lower, which is different from @fig-six results of body length and maximum length from all assemblages. It is assumed that those attributes became more standardized within the same assemblages. Compared to Suyanggae assemblages, Yongsandong has smaller CV values.

@@

>>>>

We explored assemblages that contain multiple stemmed points to examine the temporal patterns with the premise that stemmed points from one assemblage were made by the same group of people. There are four assemblages, three assemblages from the Suyanggae site (SYG) and one from the Yongsandong site (YS) that contain seven or more stemmed points. SYG1_2 and SYG6_2 are from Phase 2 and SYG1_3 and YS_3 are from Phase 3. Following the previous studies of [@bettinger1997evolutionary; @bettinger1999point; @eerkens2008cultural; @mesoudi2008cultural; @garvey2018current], we only compared CV values of attributes, body length, maximum length, mid width, and tang width. @fig-eight shows that CV values for the four assemblages are mostly below 25, which represent transmissions through indirect bias. Among the assemblages, SYG1_2 (n=7) has higher CV values as well as the widest range of confidence intervals. Compared to the Phase 2, CV values of Phase 3 are lower, which is different from @fig-plot-cv-over-time results of body length and maximum length from all assemblages. It is assumed that those attributes became more standardized within the same assemblages. Compared to Suyanggae assemblages, Yongsandong has smaller CV values.

@@

To investigate the social contexts of technological transitions represented by the emergence of stemmed points during the Korean Late Paleolithic, we applied concepts of cultural transmission theory. We asked three questions to examine the dominant mode of cultural transmission and the transmission process over time: what is the dominant mode of cultural transmission for technological innovation in the Korean Late Paleolithic? Do the modes of cultural transmission vary over space? and Do the modes of cultural transmission vary over time? We focussed on two transmission biases: guided variation (socially isolated groups made stemmed points through trial and error) and indirect bias (socially connected groups whose knowledge of stemmed points derived from copying others). We set 25 as our threshold of the CV value to decide between these two types of transmission biases, based on prior work on Great Basin Projectiles [@eerkens2001techniques] and the relatively lower quality of lithic raw materials in Korea. If the CV value is lower than 25, which indicates indirect bias while the value is 25 or higher means guided variation.

The CV values for all stemmed points are mostly over 25, except for maximum length which is 23.8. PCA analysis shows positive relationships between attributes. We also examined the CV values for chronological phases in the Korean Late Paleolithic period (@fig-plot-cv-over-time). After excluding stemmed points for the Phase 1 to compute CV due to the small sample size issue, the overall CV values become lower. In addition to the maximum length, body length, mid width, and tang width for the Phase 2 are under 25. We observe the different degree of contribution by each attribute to the shape of stemmed points. For example, tang and stem related attributes have higher CV values compared to the body related attributes. We expect that people chose to emulate the body related attributes and had more freedom to produce the tang and stem related attributes.

Compared to the  Phase 2 and 3, we found that there are only minor differences between the two phases. The CV of the two most important attributes, body length and maximum length, slightly increased in Phase 3. However the increase is not significant to generalize (p > 0.05), we explored individual assemblages including multiple stemmed points that could show a more clear pattern. The four assemblages in Suyanggae and Yongsandong sites show that CV values decreased with narrower confidence intervals, which means a high degree of precision (Figure 8<== not included yet). This could indicate the stronger connections within the groups during Phase 3.

There is no clear pattern variation in CV between the sites. Some sites such as Sachang have very high CV values while other sites including Suyanggae and Jingeunuel have lower CV (Figure 9). We assumed that the CV could be highly impacted by the choice of raw materials (Figure 10). However, similar to the CV variation for the Phase 2, the results of site variation show that CV of body length, maximum length, mid width, and tang width are generally lower than tang length, stem length, and stem width, indicating the combination of the two biases. The attributes with lower CV values could be taught through indirect bias while the higher CV attributes were depending more on various manufacturing situations. The lower CV values of the four attributes could imply that those four attributes are closely associated with the function of the tool so that the knappers wanted to standardize the shape. For example, body length is related to penetration, durability, hardness, and rejuvenation of a projectile point and tang width is connected to wound damage, penetration, durability and hardness [@odell1986experiments; @shea2001experimental; @cheshier2006projectile; @yaroshevich2016projectile; @bebber2017developing; @wood2018wound]. This combination model describes that there were social connections between groups that share the tool production technique. Individuals or groups learned the major technology of stemmed points by copying a model from another individual or group (indirect bias). Then, those individuals or groups adjusted their technique and shape of the stem depending on their needs and material availability.

# Conclusion {.unnumbered}

In this research, we examine the dominant mode of cultural transmission that might impact technological innovation in the Korean Late Paleolithic. Following the previous studies of applying cultural transmission to the introduction of bow and arrow technology in the Great Basin [@bettinger1997evolutionary; @bettinger1999point; @eerkens2008cultural; @garvey2018current], we use two transmission biases, guided variation and indirect bias to understand the transmission process and social contexts. We claim the two scenarios that explains the introduction of new technology: socially isolated groups made stemmed points through trial and error (guided variation) or socially connected groups whose knowledge of stemmed points derived from copying others (indirect bias). We ask three questions: what is the dominant mode of cultural transmission for technological innovation in the Korean Late Paleolithic? do the modes of cultural transmission vary over space? and do the modes of cultural transmission vary over time? To determine transmission biases and answer our research questions, we computed CV and PCA. 
The higher CV values (equals to or over 25) and low correlation indicate guided variation while the lower CV while the lower CV values (under 25) and high correlations represent indirect bias. Due to the small sample size, we applied confidence intervals to CV for better estimation.

Our results show that some attributes such as body length, maximum length, mid width, and tang width were transmitted through indirect bias and the other attributes were developed through guided variation. All attributes are positively related and those low variation attributes are more closely associated and highly contribute to the overall shape of the stemmed points. We assume that the guided variation played a more important role at the beginning of the transition. We observe a slight decrease in CV values over time. No clear regional patterns are confirmed. We found that CV values vary depending on raw materials. For example, coarse materials like quartzite have the highest CV values with the widest range of confidence intervals. We anticipate that our theoretical approach is applicable to other studies about technological transition and cultural transmission. We used landmarks to earn the dimensional attributes, which has potential for future research with different dimensions by calculating the distance between other sets of the existing landmarks. In our future research, we will focus on understanding actual connection between the assemblages and examining the social networks for sharing technology between the groups such as trades or physical migration to different locations.
